diffEnrich: An \proglang{R} Package to Compare Functional Enrichment Between Two Experimentally-derived Groups of Genes by Connecting to the KEGG REST API

### 2019-11-20 jss-article-rnw-diffEnrich/article.Rnw
Added text for sec 2.2 (pathEnrich) in .Rnw

>>>>

\subsection{\code{pathEnrich}: Perform enrichment analysis of individual gene sets.}

In this step, the \code{pathEnrich} function is used to identify KEGG pathways
that are enriched (i.e. over-represented) based on a gene list of interest
provided by the user. User gene lists must be ENTREZ gene IDs. If a user only
has gene symbols, the \code{clusterProfiler} package (3.9) (Yu:2012) offers a
function (\code{bitr}) that maps gene symbols and Ensembl IDs to ENTREZ gene
IDs. An example of this function's use can be found in their vignette
(\url{https://yulab-smu.github.io/clusterProfiler-book/chapter14.html#bitr}).

<<view_genelists, echo = TRUE>>=
## View sample gene lists from package data
head(geneLists$list1)
head(geneLists$list2)
@

The pathEnrich function will only use the genes from the list provided that are
also in the KEGG database. The \code{pathEnrich} function should be run at least
twice, once for the genes of interest in list 1 and once for the genes of
interest in list 2. Each \code{pathEnrich} call generates a data frame
summarizing the results of enrichment analyses in which a Fisher’s Exact test is
used to identify which KEGG pathways are enriched within the user’s list of
genes compared to all genes annotated to a KEGG pathway. Users can limit which
pathways are tested by requiring that they contain a minimum number of genes
from the list, and this can be set by changing the 'N' arguement. The default is
that a KEGG pathway must contain at least 2 genes (N = 2) from the user’s list
to be tested.

<<<pathEnrich, echo = FALSE>>=
# run pathEnrich
## List 1
list1_pe <- pathEnrich(gk_obj = kegg, gene_list = geneLists$list1, cutoff = 0.05, N = 2)
## list2
list2_pe <- pathEnrich(gk_obj = kegg, gene_list = geneLists$list2, cutoff = 0.05, N = 2)
@

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 5 results generated by \code{pathEnrich}
are in Table 2. For a detailed description of the variables in this table see Table 3.

S3 generic functions for \code{print} and \code{summary} are provided.
The \code{print} function prints the results table as a \code{tibble}, and the
\code{summary} function returns the number of pathways that reached statistical
significance as well as their descriptions, the number of genes used from the
KEGG data base, the KEGG species, and the method used for multiple testing
correction.

<<path_summary, echo = TRUE>>=
summary(list1_pe)
@

### 2019-11-20 jss-article-rnw-diffEnrich/article.tex
Added text for sec 2.2 (pathEnrich) in .Rnw

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# 3 data sets will be written as tab delimited text files}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# File location: /Users/smiharry/Documents/packages/diffEnrich}}

@@

>>>>

@@

\subsection{\code{pathEnrich}: Perform enrichment analysis of individual gene sets.}

In this step, the \code{pathEnrich} function is used to identify KEGG pathways
that are enriched (i.e. over-represented) based on a gene list of interest
provided by the user. User gene lists must be ENTREZ gene IDs. If a user only
has gene symbols, the \code{clusterProfiler} package (3.9) (Yu:2012) offers a
function (\code{bitr}) that maps gene symbols and Ensembl IDs to ENTREZ gene
IDs. An example of this function's use can be found in their vignette
(\url{https://yulab-smu.github.io/clusterProfiler-book/chapter14.html#bitr}).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## View sample gene lists from package data}
\hlkwd{head}\hlstd{(geneLists}\hlopt{$}\hlstd{list1)}
\end{alltt}
\begin{verbatim}
## [1] "361692"    "293654"    "293655"    "500974"    "100361529"
## [6] "171434"
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(geneLists}\hlopt{$}\hlstd{list2)}
\end{alltt}
\begin{verbatim}
## [1] "315547" "315548" "315549" "315550" "50938"  "58856"
\end{verbatim}
\end{kframe}
\end{knitrout}

The pathEnrich function will only use the genes from the list provided that are
also in the KEGG database. The \code{pathEnrich} function should be run at least
twice, once for the genes of interest in list 1 and once for the genes of
interest in list 2. Each \code{pathEnrich} call generates a data frame
summarizing the results of enrichment analyses in which a Fisher’s Exact test is
used to identify which KEGG pathways are enriched within the user’s list of
genes compared to all genes annotated to a KEGG pathway. Users can limit which
pathways are tested by requiring that they contain a minimum number of genes
from the list, and this can be set by changing the 'N' arguement. The default is
that a KEGG pathway must contain at least 2 genes (N = 2) from the user’s list
to be tested.

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 5 results generated by \code{pathEnrich}
are in Table 2. For a detailed description of the variables in this table see Table 3.

S3 generic functions for \code{print} and \code{summary} are provided.
The \code{print} function prints the results table as a \code{tibble}, and the
\code{summary} function returns the number of pathways that reached statistical
significance as well as their descriptions, the number of genes used from the
KEGG data base, the KEGG species, and the method used for multiple testing
correction.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{summary}\hlstd{(list1_pe)}
\end{alltt}
\begin{verbatim}
## 219 KEGG pathways were tested. 
##  Only KEGG pathways that contained at least 2 genes from gene_list were tested. 
##  KEGG pathway species: Rattus norvegicus (rat)
##  8856 genes from gene_list were in the KEGG data pull. 
##  p-value adjustment method: BH
##  36 pathways reached statistical significance after multiple testing correction at a cutoff of 0.05. 
##  
## Significant pathways: 
##  Tight junction
## Yersinia infection
## Colorectal cancer
## Choline metabolism in cancer
## Endometrial cancer
## Endocytosis
## Neurotrophin signaling pathway
## Thermogenesis
## Oocyte meiosis
## VEGF signaling pathway
## Thyroid hormone signaling pathway
## Hippo signaling pathway
## T cell receptor signaling pathway
## Apoptosis
## Hepatocellular carcinoma
## MAPK signaling pathway
## Focal adhesion
## Salmonella infection
## Non-alcoholic fatty liver disease (NAFLD)
## ErbB signaling pathway
## Sphingolipid signaling pathway
## Pancreatic cancer
## Progesterone-mediated oocyte maturation
## Alzheimer disease
## Endocrine resistance
## Adrenergic signaling in cardiomyocytes
## IL-17 signaling pathway
## Chronic myeloid leukemia
## Dopaminergic synapse
## Prostate cancer
## EGFR tyrosine kinase inhibitor resistance
## Hepatitis C
## Ras signaling pathway
## Acute myeloid leukemia
## Insulin signaling pathway
## Fc epsilon RI signaling pathway
\end{verbatim}
\end{kframe}
\end{knitrout}
